Long-term trends in the contribution of major causes of death to the black-white life expectancy gap by US state. A retrospective study of all deaths between 1969 and 2013

### 2017-06-28 Manuscript/Manuscript_COD_Trends_BMJOpen.Rmd
structured abstract

---
title: Long-term trends in the contribution of major causes of death to the black-white life expectancy gap by US state
bibliography: bibliography.bib
output:
  word_document: default
  html_document: default
always_allow_html: yes
---

@@

States have fared differently in their progress towards eliminating the black-white life expectancy gap. Our objective is to describe the pattern of contributions of six major causes of deaths to the sex-specific black-white life expectancy gap across states over the last half-century, and identify divergent states.

Using vital statistics and census data, we extracted the number of deaths and population sizes for the years 1969 to 2013, by state, gender, race, 19 age groups, and six major causes of death.

@@

Although mortality from cardiovascular disease has decreased dramatically, its contribution to the life expectancy gap increased over time for men (from `r cvd.male.1969` to `r cvd.male.2013` years), but decreased for women (from `r cvd.female.1969` to `r formatC(as.numeric(cvd.female.2013), digits = 1)` years). The contribution of non-communicable diseases to the gap was stable over time for men (approximately `r ncd.male.average` years) but decreased for women (from `r ncd.female.1969` to `r ncd.female.2013` years), while cancers exhibited an inverted-U trend for men (peaking at `r can.male.max` years in 1988) and a stable contribution for women (approximately `r can.female.average` years). Both genders exhibited a decreased contribution from injuries (men: `r inj.male.1969` to `r inj.male.2013` years), that became negative for women (women: `r inj.female.1969` to `r inj.female.2013` years). Several states diverged from these general trends.

Life expectancy for both races has improved substantially in the US. For men, much of this improvement was due to narrowing differences in injury-related mortality, but these contributions were rivalled by an increasing gap in CVD-related mortality. In women, a crossover in injury-related mortality led to a narrower gap, realized partially by increasing mortality among whites.

>>>>

---
title: Long-term trends in the contribution of major causes of death to the black-white life expectancy gap by US state. A retrospective study of all deaths between 1969 and 2013
bibliography: bibliography.bib
output:
  word_document: default
  html_document: default
always_allow_html: yes
---

@@

@@

Objectives: Our objective is to describe the pattern of contributions of six major causes of deaths to the sex-specific black-white life expectancy gap across states over the last half-century, and identify divergent states.

Design: A repeated cross-sectional study of all deaths

Setting: 39 US States and the District of Columbia

Participants: Using vital statistics and census data, we extracted the number of deaths and population sizes for the years 1969 to 2013, by state, gender, race, 19 age groups, and six major causes of death.

Primary and Secondary outcome measures: We estimated the contribution (in years) that each of six major causes of death to the gap in life expectancy between black and white individuals.

Results: Although mortality from cardiovascular disease has decreased dramatically, its contribution to the life expectancy gap increased over time for men (from `r cvd.male.1969` to `r cvd.male.2013` years), but decreased for women (from `r cvd.female.1969` to `r formatC(as.numeric(cvd.female.2013), digits = 1)` years). The contribution of non-communicable diseases to the gap was stable over time for men (approximately `r ncd.male.average` years) but decreased for women (from `r ncd.female.1969` to `r ncd.female.2013` years), while cancers exhibited an inverted-U trend for men (peaking at `r can.male.max` years in 1988) and a stable contribution for women (approximately `r can.female.average` years). Both genders exhibited a decreased contribution from injuries (men: `r inj.male.1969` to `r inj.male.2013` years), that became negative for women (women: `r inj.female.1969` to `r inj.female.2013` years). Several states diverged from these general trends.

Conclusions: Life expectancy for both races has improved substantially in the US. For men, much of this improvement was due to narrowing differences in injury-related mortality, but these contributions were rivalled by an increasing gap in CVD-related mortality. In women, a crossover in injury-related mortality led to a narrower gap, realized partially by increasing mortality among whites.
